|

### 2024-08-26 reproducibility/manuscript/manuscript.qmd
fix(manuscript): remove paragraph break in align env
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
latex fixes and updates to methods. ready to be read.

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
\end{align}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}, \label{eq-dsdt}.
\end{align}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown activation state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times $\tau_{cg}$, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are in fact unknown.

@@

First, we introduce the generative model that describes the various unobserved times:
\begin{align}
  % unit lognormal t_c
  t_c &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_0
  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
  % switching time
  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_1
  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
  %cell-gene-specific activation state
  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
  % cell-gene-specific latent time
  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
\end{align}
Here, $\tau_{cg}$ represents the displacement of time per
cell and gene with 
\begin{align}
 \text{softplus}(t) :=  \log( 1 + e^t).
\end{align}
Recall that $t_c$ is the shared time per cell, $t^{(k_{cg})}_{0,g}$ is the
gene-specific switching time. Each cell and gene combination has its
transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
activation state and $1$ indicates the expression state. Each gene has two
switching times for representing activation and repression: $t^{(0)}_{0,g}$ is
the first switching time corresponding to when the gene expression starts to be
activated, $t^{(1)}_{0,g}$  is the second switching time corresponding to when
the gene expression starts to be repressed.

@@

Now, we describe the priors for the initial conditions, noting that this is the only difference between Model 1 and Model 2:
\begin{align}
  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
    (0, 0) & \text{Model 1} \\
    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
  \end{cases}

  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
  \end{cases}
\end{align}

Finally, we describe the ODE solution at time $\tau_{cg}$, and the observation model at those times that give rise to the observed counts:
\begin{align}
  \hat{u}_{cg}, \hat{s}_{cg} &= \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big) \\
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_cg, \hat{s}_cg$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}

>>>>

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
\end{align}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
\end{align}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown activation state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times $\tau_{cg}$, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are in fact unknown.

@@

First, we introduce the generative model that describes the various unobserved times:
\begin{align}
  % unit lognormal t_c
  t_c &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_0
  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
  % switching time
  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_1
  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
  %cell-gene-specific activation state
  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
  % cell-gene-specific latent time
  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
\end{align}
Here, $\tau_{cg}$ represents the displacement of time per
cell and gene with 
\begin{align}
 \text{softplus}(t) :=  \log( 1 + e^t).
\end{align}
Recall that $t_c$ is the shared time per cell, $t^{(k_{cg})}_{0,g}$ is the
gene-specific switching time. Each cell and gene combination has its
transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
activation state and $1$ indicates the expression state. Each gene has two
switching times for representing activation and repression: $t^{(0)}_{0,g}$ is
the first switching time corresponding to when the gene expression starts to be
activated, $t^{(1)}_{0,g}$ is the second switching time corresponding to when
the gene expression starts to be repressed, and is determined by the first
switching time and the gene-specific switching time $\Delta \text{switching}_g$.
The cell-gene-specific activation state $k_{cg}$ is a Bernoulli random variable
with logits equal to the difference between the cell's shared time $t_c$ and the time $t^{(1)}_{0,g}$ when the gene expression starts to be repressed.

@@

Now, we describe the priors for the initial conditions, noting that this is the only difference between Model 1 and Model 2:
\begin{align}
  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
    (0, 0) & \text{Model 1} \\
    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
  \end{cases} \\
  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
  \end{cases}
\end{align}

We define the ODE solution at time $\tau_{cg}$ as:
\begin{equation}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed counts as:
\begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}
